\documentclass{article}

% --- Packages ---
\usepackage[utf8]{inputenc} % Allows for special characters
\usepackage{amsmath}        % For math symbols
\usepackage{graphicx}       % For adding images later
\usepackage{float}

% --- Document Info ---
\title{SAiDL Assignment 2026}
\author{Joel Thomas}
\date{27th March 2026 Onwards}

\begin{document}

\maketitle
Core-ML Task
\section{Intro}
This is my report on the Core-ML task of Attention-based Transformer, trained on WikiText-2. I'll be updating my progress here.

\section{Basic Transformer Baseline}
From the 'Attention is All You Need' article, the core of the Transformer is the Scaled Dot-Product Attention:

\begin{equation}
    \text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)V
\end{equation}

\noindent where:
\begin{itemize}
    \item $Q$ $\rightarrow$ Queries
    \item $K$ $\rightarrow$ Keys
    \item $V$ $\rightarrow$ Values
    \item $d_k$ $\rightarrow$ Dimension of the keys
\end{itemize}

\section{Baseline Model Transformer Details}
I've used AI to explain what each class does in the Transformer code.

\subsection{Data Processing and Embeddings}
\textbf{InputEmbedding}: Converts input tokens into dense vectors of dimension $d_{model}$ and scales them by $\sqrt{d_{model}}$ to stabilize gradients. It maps the discrete vocabulary to a continuous space suitable for neural network processing.

\medskip % Adds a nice small gap
\textbf{PositionalEncoding}: Injects information about the relative or absolute position of tokens using sine and cosine functions of different frequencies. This allows the model to utilize the order of the sequence, as the self-attention mechanism itself is permutation-invariant.

\subsection{Model Components}
\textbf{LayerNormalization}: Standardizes the inputs across the features dimension to a mean of zero and variance of one, followed by a learnable affine transformation. This ensures stable training and faster convergence by mitigating internal covariate shift.

\medskip
\textbf{FeedForwardBlock}: Consists of two linear transformations with a ReLU activation and dropout in between to process each position independently. It increases the model's capacity to learn complex patterns within the hidden representations.

\medskip
\textbf{MultiHeadAttentionBlock}: Projects the $Q, K,$ and $V$ into multiple subspaces to attend to information from different representation aspects simultaneously. It uses the previously-mentioned scaled dot-product attention in parallel across $h$ heads before concatenating and re-projecting the results.

\subsection{Parameters Used for WikiText-2 Training}

The baseline Transformer was trained on the WikiText-2 dataset using the following configuration:

\begin{table}[H]
\centering
\caption{Hyperparameters for Baseline Transformer Training}
\begin{tabular}{lll}
\hline
\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\
\hline
\multicolumn{3}{l}{\textit{Model Architecture}} \\
\texttt{d\_model} & 128 & Embedding dimension \\
\texttt{n\_heads} & 4 & Number of attention heads \\
\texttt{n\_layers} & 4 & Number of decoder layers \\
\texttt{d\_ff} & 512 & Feed-forward hidden dimension \\
\texttt{dropout} & 0.1 & Dropout rate \\
\hline
\multicolumn{3}{l}{\textit{Sequence}} \\
\texttt{seq\_len} & 1024 & Context window length (tokens) \\
\texttt{vocab\_size} & 50{,}257 & GPT-2 BPE tokenizer vocabulary size \\
\hline
\multicolumn{3}{l}{\textit{Training}} \\
\texttt{batch\_size} & 8 & Sequences per batch \\
\texttt{grad\_accum\_steps} & 4 & Gradient accumulation steps \\
\texttt{effective\_batch} & 32 & Effective batch size (batch $\times$ accum) \\
\texttt{lr} & $3 \times 10^{-4}$ & Peak learning rate (AdamW) \\
\texttt{weight\_decay} & 0.1 & AdamW weight decay \\
\texttt{warmup\_steps} & 200 & Linear warmup steps \\
\texttt{max\_steps} & 5{,}000 & Total training steps \\
\texttt{eval\_interval} & 250 & Steps between validation evaluations \\
\texttt{eval\_iters} & 50 & Batches averaged per validation estimate \\
\hline
\end{tabular}
\label{tab:hyperparams}
\end{table}

\subsection{Results}

Table \ref{tab:results} summarizes the training and validation metrics recorded 
every 250 steps over the full 5{,}000-step training run on WikiText-2.

\begin{table}[h]
\centering
\caption{Training and Validation Metrics Over 5{,}000 Steps}
\begin{tabular}{rrrrrr}
\hline
\textbf{Step} & \textbf{Train Loss} & \textbf{Train PPL} & \textbf{Val Loss} & \textbf{Val PPL} & \textbf{Tok/s} \\
\hline
250  & 7.44 & 1696.88 & 7.47 & 1759.35 & 29{,}580 \\
500  & 6.60 & 735.60  & 6.71 & 817.41  & 29{,}258 \\
750  & 6.08 & 436.26  & 6.23 & 508.59  & 29{,}118 \\
1000 & 5.76 & 316.06  & 6.03 & 417.27  & 29{,}117 \\
1250 & 5.53 & 251.85  & 5.87 & 353.61  & 29{,}033 \\
1500 & 5.33 & 206.69  & 5.77 & 319.71  & 29{,}094 \\
1750 & 5.15 & 172.62  & 5.66 & 288.23  & 29{,}044 \\
2000 & 5.03 & 153.57  & 5.62 & 276.00  & 29{,}040 \\
2250 & 4.92 & 137.67  & 5.57 & 261.15  & 29{,}130 \\
2500 & 4.86 & 129.11  & 5.54 & 255.69  & 29{,}052 \\
2750 & 4.73 & 112.97  & 5.48 & 239.11  & 29{,}046 \\
3000 & 4.70 & 109.93  & 5.48 & 240.87  & 29{,}107 \\
3250 & 4.64 & 103.94  & 5.45 & 232.15  & 29{,}115 \\
3500 & 4.61 & 100.23  & 5.44 & 231.49  & 29{,}081 \\
3750 & 4.56 & 95.45   & 5.45 & 232.93  & 29{,}015 \\
4000 & 4.56 & 96.05   & 5.44 & 230.93  & 29{,}011 \\
4250 & 4.54 & 94.15   & 5.43 & 229.01  & 29{,}072 \\
4500 & 4.56 & 95.62   & 5.45 & 232.38  & 29{,}101 \\
4750 & 4.51 & 90.87   & 5.45 & 231.87  & 29{,}111 \\
5000 & 4.55 & 95.02   & 5.41 & \textbf{224.70}  & 29{,}034 \\
\hline
\end{tabular}
\label{tab:results}
\end{table}

The model converged to a best validation perplexity of \textbf{224.70} at step 
5{,}000, with a corresponding training perplexity of 95.02. The gap between 
training and validation perplexity indicates mild overfitting, which is expected 
given the relatively small model size (13.71M parameters) and limited dataset. 
Throughput remained highly stable throughout training at approximately 
\textbf{29{,}000 tokens/sec} on an NVIDIA Tesla T4 GPU, demonstrating consistent 
hardware utilization. Validation perplexity improved rapidly in the early stages 
(dropping from 1759 to 255 in the first 2{,}500 steps) before plateauing around 
229--232 for the final 1{,}500 steps, suggesting the model had largely converged 
by step 3{,}500.


\end{document}
